---
title: "STAT331 Project"
author: "Cassidy Gertzfield, Brandon Montalvo, Lauren Hu, Rapha Coutin"
format:
  html:
    embed-resources: true
    code-tools:       true
    toc:              true
    code-fold:        true
editor: source
execute:
  error:   true
  echo:    true
  message: false
  warning: false
---

# Happiness and Income Analysis

```{r setup, include = TRUE, echo = FALSE}
# load packages
library(readr)
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
library(gganimate)
library(gifski)
```

```{r importdata, include = TRUE, echo = FALSE}
# import data sets
happy <- read_csv("hapiscore_whr.csv")
income <- read_csv("mincpcap_cppp.csv")
```

```{r pivot,include = TRUE, echo = FALSE}
## make long data set 
happy <- happy|>
  pivot_longer(cols = `2005`:`2023`,
               names_to = "year",
               values_to = "happy_score")

income <- income|>
  pivot_longer(cols = `1800`:`2100`,
               names_to = "year",
               values_to = "avg_daily_inc")
```

```{r merge, include = TRUE, echo = FALSE}
## merge into one data set 
happy_income <- happy|>
  left_join(income, 
            by = c("year","country"))
```

```{r NAs, include = TRUE, echo = FALSE}
## deal with missing values 
happy_income <- happy_income|>
  filter(!is.na(happy_score) & !is.na(avg_daily_inc))
```

## Introduction & Data

This analysis explores the relationship between average household income and life satisfaction across countries over time.

The primary data on life satisfaction comes from the World Happiness Report published by the Sustainable Development Solutions Network, based on surveys conducted by the Gallup World Poll. The key question, known as the Cantril Ladder, asks people to imagine a ladder from 0 (worst possible life) to 10 (best possible life) and say where they stand right now. Gapminder imported this data for 149 counties from 2005-2020. For recent years, about two-thirds of countries have 2020 survey data and the rest use 2018 or 2019 data.

Income data is compiled by Gapminder, which integrates World Bank household survey data and historical estimates. The income data shows the average daily household income or spending per person, adjusted to 2017 purchasing power parity (PPP) dollars. For countries lacking direct survey data, Gapminder estimated income based on GDP per capita and extended estimates historically back to 1800 and forward to 2100 using GDP growth rates.

Our observational unit is a combination of year and country. Our data set includes average household income and life satisfaction scores for up to 149 countries, mostly between 2005 and 2020. The data for happiness score collected by Gallup World Poll is reported in the World Happiness Report and is put together by Gapminder (happiness score was converted from a 0-10 scale to a 0-100 scale), and the data for daily income (for years 1981 to 2019) comes from the World Development Indicators published by the World Bank and is compiled by Gapminder.

We hypothesize that a country’s average daily income and happiness are positively related, with happiness increasing more per increase in average income for smaller average incomes and tapering off towards the upper end of income values. On an individual level, the argument could be made that having more money feels hollow and leads to a decrease in happiness. However, we are looking at the average household income and happiness score across entire countries. A higher average daily income may relate to lower poverty levels and more disposable income, and thus less stress. A report from [outworldindata.com](http://outworldindata.com) (linked below) shows a clear positive relation between GDP per capita and average happiness (as registered on the Cantril Ladder scale) across multiple countries. The assumption is that when a country’s people are able to meet their basic needs and have financial security, they are happier.

### Cleaning Process

To prepare our data for analysis, we started with two data sets—one with world happiness scores and one with average daily income—and reorganized both so that each row represents a specific country and year. We then combined the two datasets so that each row included both the happiness score and income data for that country and year. If either the happiness score or income value was missing for a given country and year, we removed that row to keep the analysis consistent and reliable. After this cleaning process, we had 2,342 complete records covering countries from the years 2005 to 2023.

## Linear Regression

Linear regression is a modeling method to evaluate the relationship between an explanatory outcome variable and one or more explanatory predictor variables. Linear regression uses the ordinary least squares method (OLS) to get estimated values that make the sum of squared errors (actual - predicted values) as small as possible. This method allows for an evaluation of the direction and strength of the predictor's effect in the model. There are four main assumptions to run a linear regression: linearity, independence, constant variance of residuals, and normally distributed residuals.

In order to conduct the following linear regression, for each country across all available years, we took the averages of their happiness score and average daily income. This made it so we can have one value for each country in the data set. Additionally, we transformed our explanatory variable (average daily income) with a natural log to better fit the data with a linear model.

### Visualizations

For preliminary analysis, the following scatter plot (Figure 1) shows the relationship between average daily income and happiness score. Each point represents a country's averages of happiness score and average daily income across all years.

```{r}
#| fig-cap: "Figure 1: Scatter plot without Transformation"

happy_income |> 
  group_by(country) |> 
  summarise(happy_score = mean(happy_score),
            avg_daily_inc = mean(avg_daily_inc)) |> 
  ggplot(mapping = aes(x = avg_daily_inc, 
                       y = happy_score)) + 
  geom_point() + 
  labs(title = "Countries' Happiness Score and Average Daily Income", 
       subtitle = "Happiness Score (0 - 100)", 
       x = "Average Daily Income (2017 PPP Dollars)", 
       y = "", 
       caption = "Happiness score and average daily income are averaged across all available years (2005 - 2023) for each country in the data set") + 
  theme_bw()
```

As we can see, the relationship between average daily income and happiness is positive, but the form is not linear so a transformation will be needed to perform a linear analysis.

The following plot displays how the trend between the average daily income and happiness score per country changes over the years 2005-2023 without a log transformation.

```{r}
#| fig-cap: "Figure 2: Animated - Relationship Between Happiness Score and Avg Daily Income Over Time"

anim_plot <- happy_income |> 
  mutate(year = as.integer(year)) |> 
  ggplot(mapping = aes(x = avg_daily_inc, 
                       y = happy_score, 
                       size = 10)) + 
  geom_point(alpha = 0.5,
             color = "steelblue4",
             show.legend = F) + 
  labs(x = "Average Daily Income (2017 PPP Dollars)", 
       y = "", 
       title = "Country's Happiness Score and Average Daily Income Years 2005 - 2023", 
       subtitle = "Happiness Score (0 - 100)                    Year: {frame_time}") + 
  theme_bw() +
  transition_time(year) + 
  enter_fade() + 
  exit_reset() + 
  ease_aes('cubic-in-out')

animate(anim_plot, 
        renderer = gifski_renderer(), 
        nframes = 500, fps = 25)
```

According to Figure 2, the positive curvilinear relationship between the average daily income and happiness score remains relatively consistent throughout 2005 to 2023.

The following scatter plot displays the relationship between the happiness score and the log of average daily income by country.

```{r LM}
#| fig-cap: "Figure 3: Linear Regression Plot"

# getting the mean of both variables for each country
happy_income_averaged <- happy_income |> 
  group_by(country) |> 
  summarise(happy_score = mean(happy_score),
            avg_daily_inc = log(mean(avg_daily_inc))) 

# plotting the linear regression with a scatter plot
happy_income_averaged |> 
  ggplot(mapping = aes(x = avg_daily_inc,
                       y = happy_score)) + 
  geom_point() +
  geom_smooth(method = "lm") + 
  labs(title = "Happiness vs Average Daily Income with Log transformation",
       subtitle = "Happiness Score",
       x = "log(Average Daily Income)",
       y = "",
       caption = "Happiness score and average daily income are averaged across all available years for each country in the data set.") + 
  theme_bw()
```

According to Figure 3, there is a strong, positive linear relationship with no apparent unusual values between a country's happiness score and their average daily income (after a natural log transformation on average daily income). This means that as average daily income increases in a country, the estimated happiness score is also expected to increase.

### Regression Model Estimates

The following table displays the model estimates for the linear regression ran above.

```{r estimates}
#| tbl-cap: "Table 1: Regression Model Estimates"

# Creating the linear model
model <- lm(happy_score ~ log(avg_daily_inc),
            data = happy_income_averaged) 


tidy(model) |> 
  kable(col.names = c("Term",
                      "Estimate",
                      "Standard Error",
                      "Statistic",
                      "P-Value"))|>
  kable_styling(position = "center",
                bootstrap_options = "striped")|>
  row_spec(0, bold = T) |> 
  add_header_above(header = c("Linear Model Coefficients" = 5), bold = TRUE)

```

According to Table 1, the intercept for the model is 38.24. This means that when a country's average daily income per household is \$1, their predicted average happiness score is 38.24 out of 100. The coefficient for the log-average daily income is 18.62. As we used a natural log transformation on average daily income, this means that if we multiply the average daily income by $e$ (approximately 2.72), their predicted happiness score increases by 18.62 points.

The linear model with the transformation is: $$ \hat{y} = 38.24 + 18.62\ln(x_i),$$ where $\hat{y}$ is the estimated happiness score out of 100 and $x_i$ is a country's average daily income in 2017 purchasing power parity (PPP) dollars.

If we wanted to represent the model as a true linear model, we would write it as $$ \hat{y} = 38.24 + 18.62 \cdot z_i,$$ where $\hat{y}$ is the estimated happiness score out of 100 and $z_i$ is the natural log of the average daily income in 2017 purchasing power parity (PPP) dollars per country. However, for the sake of this report, we will be using the model given in ($1$).

### Model Fit Estimates

The following table displays the statistics for the fit of the regression model.

```{r modelfit}
#| tbl-cap: "Table 2: Model Fit Estimates"

var_response <- var(happy_income_averaged$happy_score)
var_fitted <- var(fitted(model))
var_residuals <- var(residuals(model))
r_squared <- var_fitted / var_response

model_fit_summary <- tibble(
  Statistic = c("Variance in Response (A)", 
                "Variance in Fitted Values (B)", 
                "Variance in Residuals", 
                "Model R-squared (B/A)"),
  Value = c(var_response, 
            var_fitted, 
            var_residuals, 
            r_squared)
)

model_fit_summary |>
  kable(digits = 3, caption = "Model Fit Summary") |>
  kable_styling(bootstrap_options = "striped")
```

According to Table 2, the R-squared value of 0.658 indicates that approximately 65.8% of the variability in country's happiness scores is explained by the regression model containing log of average daily income, indicating that the model containing log of average daily income to predict happiness is a moderate fit.

## Cross Validation: *K* Folds

This section conducts a k-fold cross validation method. This validation is important because it gives a sense of how well our model can be used to predict new data, rather than simply fitting the data it has already been tested on. The first step is to have the user decide on the number of folds, k. In this case, we have 14 k folds which allows us to have at least ten observations in each fold. Next, we must split the data randomly into the 14 groups. In each group, we will fit the same regression model as above and predict the outcomes to calculate how much of the variation in happiness score is explained by average daily income (model R\^2). We repeat this process for all 14 folds and take the average. We then have to compare the original model's R\^2 with the average R\^2 to see whether the model is over or under fitting. In the case of over-fitting, the model R\^2 will perform better (be higher) than the average R\^2. The case of under-fitting would arise if the model R\^2 is worse (is lower) than the average R\^2. If the R\^2 values are relatively equal, the model is neither over nor under-fit.

```{r kfold}

k <- 14
n <- 163


happy_income_averaged <- happy_income_averaged |> 
  mutate(fold_random = sample(rep_len(1:k, length.out = n),
                       size = n)) 


get_fold_r2 <- function(fold_number, data) {
  test_data  <- data |> filter(fold_random == fold_number)

  model <- lm(happy_score ~ log(avg_daily_inc), 
              data = data,
              subset = fold_random != fold_number)

  predictions <- predict(model, newdata = test_data)

  var_predicted <- var(predictions)
  var_actual    <- var(test_data$happy_score)

  r2 <- var_predicted / var_actual
  return(r2)
}

r2_results <- tibble(
  fold = 1:k,
  r2   = map_dbl(1:k, ~ get_fold_r2(.x, happy_income_averaged))
)

mean_r2 <- mean(r2_results$r2)

mean_r2|>
  kable(digits = 3,
        col.names = "Mean R²",
        caption = "Mean R² Value for K-Folds")|>
  kable_classic(full_width = F,
                bootstrap_options = "striped") |> 
  row_spec(row = 0, bold = T, align = "c")
```

### K Fold Plot

```{r kplot}
r2_results|>
  ggplot(aes(x = fold,
             y = r2)) + 
  geom_point() + 
  geom_hline(yintercept = mean_r2, color = "orange3", linetype = 2) + 
  annotate(
    "text",
    x      = Inf,                
    y      = mean_r2,           
    label  = "Mean R² Across Folds",
    color  = "orange3",
    hjust  = 1.1,                
    vjust  = -0.5                
  ) + 
  labs(
    x = "Fold Number",
    y = "",
    title = "Plot of R² Values for K Fold Cross Validation",
    subtitle = "R² Value") + 
  theme_bw()
```

## Conclusion

In summary, this analysis shows a positive association between average daily income and happiness score across countries. Our linear regression showed predictive performance **INSERT K FOLD CONCLUSION.** Our findings suggest that while income may not be the sole determinant of happiness, it remains a significant predictor. This underscores the value of economic development alongside social policies for a nation's happiness. In the future, we suggest adding more control variables and checking model conditions more thoroughly to have a more accurate gauge of the results.

## References

Ortiz-Ospina, E., & Roser, M. (2024). *Happiness and life satisfaction*. Our World in Data. <https://ourworldindata.org/happiness-and-life-satisfaction>

GitHub was used for collaboration on this report. The GitHub repository can be accessed at: <https://github.com/montalvoB/STAT-331-Project>
